\documentclass[10pt,twocolumn]{article}

\usepackage[margin=0.75in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage[numbers]{natbib}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\newcommand{\GroupID}{GROUP\_ID}
\newcommand{\PrimaryModel}{Qwen3-4B-Instruct-2507}
\newcommand{\placeholder}[1]{\textcolor{red}{[TBD: #1]}}

\setlength{\columnsep}{0.22in}
\setlist[itemize]{leftmargin=*,topsep=2pt,itemsep=1pt}
\setlist[enumerate]{leftmargin=*,topsep=2pt,itemsep=1pt}

\title{A Lightweight Verifier-Based Clarification Guard for Under-Specified Tool Calls on BFCL\\
\large Fixed-Budget Robustness for a Small Open Model}
\author{Group \GroupID\\
University of Edinburgh\\
\texttt{\placeholder{student numbers / names if required}}}
\date{}

\begin{document}

\maketitle

\begin{abstract}
Tool use is now a standard capability for LLM agents, but benchmark results continue to show that small open models remain brittle when requests are under-specified or require stateful multi-turn interaction. This project studies a narrow empirical question on the Berkeley Function-Calling Leaderboard (BFCL): can a lightweight inference-time clarification guard reduce harmful tool calls under a fixed interaction budget? We focus on a single 4B open model, \PrimaryModel{}, and analyse the gap between normal multi-turn function calling and a harder missing-parameter setting. The official direct baseline already shows a substantial drop from 27.50\% on \texttt{multi\_turn\_base} to 18.00\% on \texttt{multi\_turn\_miss\_param}, motivating a targeted intervention. The proposed method does not fine-tune the model. Instead, it first generates a draft tool call, then asks the same model to verify whether the draft should be executed or replaced by a clarification question. On a 200-turn pilot exported from the official \texttt{multi\_turn\_miss\_param} split, this verifier-based guard reduces harmful tool calls on empty-ground-truth turns from 66.67\% to 14.58\%, but it also over-clarifies many required-action turns, increasing the no-call rate on required turns from 22.37\% to 59.21\%. The contribution is therefore a focused, reproducible systems study of a concrete robustness--latency trade-off rather than a new learning algorithm.
\end{abstract}

\section{Introduction}
Tool use extends language models beyond free-form text generation and is now central to assistant-style systems. Toolformer showed that language models can learn to call APIs and incorporate their outputs into generation \citep{schick2023toolformer}. ReAct later demonstrated the broader principle of interleaving reasoning with actions for interactive tasks \citep{yao2023react}. These ideas motivate modern tool-calling assistants, but they do not by themselves guarantee reliable behaviour when the task requires structured state tracking, missing-parameter handling, or recovery from invalid actions.

The Berkeley Function-Calling Leaderboard (BFCL) was introduced to evaluate exactly these behaviours in a standardised way \citep{patil2025bfcl}. Its agentic and multi-turn categories show that tool use remains brittle even when the model performs well on simpler one-shot calls \citep{patil2025bfcl}. Recent analyses reinforce the same point from different perspectives. FAIL-TaLMs identifies under-specified requests and unavailable tools as recurring failure modes \citep{trevino2025failtalms}. STYLE and AskToAct study clarification behaviour in conversational settings \citep{chen2024style,zhang2025asktoact}. HiTEC and CRITICTOOL focus on parameter-checking and tool-use self-critique \citep{cui2025hitec,huang2025critictool}. Together, these results suggest that deciding \emph{when not to act immediately} is a real systems problem rather than a prompt-formatting detail.

This coursework project focuses on a deliberately narrow version of that problem. Rather than training a new model or collecting additional preference data, we ask whether a lightweight \emph{verifier-based clarification guard} can improve robustness for a small open model under a fixed interaction budget. The central observation is simple: the direct BFCL baseline for \PrimaryModel{} is much worse on the missing-parameter subset than on the normal multi-turn subset. If that drop is driven by premature or harmful tool calls, then a small amount of extra control at inference time may help even without any weight updates.

The contribution is therefore empirical and system-oriented. We study a test-time policy that first asks the model for a draft tool call and then runs a second-pass verifier prompt to decide whether the draft should be executed or replaced by a clarification question. We then compare this policy to the official direct baseline and to a slightly broader escalation policy that preserves direct behaviour on normal cases. This is a modest claim, but it is appropriate for a reproducible coursework project: the goal is not to claim a new foundation-model capability, but to analyse one concrete failure mode with a controlled intervention and standard metrics.

\section{Task and Data}
\subsection{Research Question}
The main research question is:
\begin{quote}
\emph{Can a lightweight verifier-based clarification guard reduce harmful tool calls on under-specified BFCL multi-turn cases for a small open model under a fixed inference budget, and what robustness--latency trade-off does it introduce?}
\end{quote}

The claim is intentionally narrow. We do not attempt to improve all BFCL categories equally, and we do not claim to invent clarification or repair from scratch. Instead, we study a focused inference-time control policy on the subset where it is most likely to matter.

\subsection{Benchmark and Subsets}
The project uses the official BFCL pipeline and evaluator \citep{patil2025bfcl}. BFCL is well suited to this coursework because it provides a current tool-calling benchmark, official scripts, and agentic subsets that stress the exact behaviours of interest.

The primary subsets used in this project are:
\begin{itemize}
    \item \texttt{multi\_turn\_base}: normal multi-turn tool calling;
    \item \texttt{multi\_turn\_miss\_param}: under-specified multi-turn requests with missing arguments;
    \item \texttt{memory\_kv}: a lightweight negative-control slice to check whether any gains are specific to missing-parameter handling rather than generic across all agentic tasks.
\end{itemize}

The main analysis is centred on \texttt{multi\_turn\_miss\_param}. The other subsets are used to anchor the interpretation: \texttt{multi\_turn\_base} tests whether the intervention hurts normal multi-turn behaviour, and \texttt{memory\_kv} tests whether the same policy transfers to a different failure mode where clarification is less obviously useful.

\subsection{Model and Serving Setup}
The primary model is \PrimaryModel{} \citep{yang2025qwen3,qwen2025modelcard}, served locally through an OpenAI-compatible \texttt{vLLM} endpoint. The local serving configuration uses host \texttt{127.0.0.1}, port \texttt{1053}, \texttt{dtype=auto}, \texttt{max\_model\_len=32768}, and \texttt{gpu\_memory\_utilization=0.9}. For the local policy runner, decoding is deterministic with temperature $0.0$ and a maximum of $384$ output tokens. All experiments in this report were run on the local coursework machine with an RTX 5090 Laptop GPU, 64 GB RAM, and an Intel U9 275H CPU. This setup matches the practical scope of the coursework: a small open model that can be evaluated reproducibly on a single local GPU without depending on cluster scheduling for the main results.

\subsection{Current Direct Baseline}
The official direct baseline has already been run on the selected BFCL slices. The current category-level results are shown in Table~\ref{tab:baseline}.

\begin{table}[t]
\centering
\small
\begin{tabular}{lc}
\toprule
Subset & Direct Accuracy \\
\midrule
\texttt{multi\_turn\_base} & 27.50\% \\
\texttt{multi\_turn\_miss\_param} & 18.00\% \\
\texttt{memory\_kv} & 12.26\% \\
\texttt{memory\_vector} & 16.13\% \\
\texttt{memory\_rec\_sum} & 27.74\% \\
\bottomrule
\end{tabular}
\caption{Current official BFCL direct baseline for \PrimaryModel{} on the evaluated subsets.}
\label{tab:baseline}
\end{table}

The most important observation is the drop from 27.50\% to 18.00\% between \texttt{multi\_turn\_base} and \texttt{multi\_turn\_miss\_param}. This provides a concrete motivation for the proposed method: under-specified requests are already a demonstrated weakness for the direct baseline.

\section{Methodology}
\subsection{Baseline System}
The direct baseline uses the official BFCL generation and evaluation pipeline with no extra controller. Given a user request, dialogue history, and tool schema, the model directly emits tool calls or natural-language fallback text. This setting is important because it establishes a standard reference point and keeps the evaluator unchanged.

\subsection{Clarification Guard}
The main proposed method is a lightweight \textbf{Clarify-Guard}. It is an inference-time controller that operates before any tool call is executed. The guard performs two steps:
\begin{enumerate}
    \item Ask the model for a \emph{draft} tool call under the standard direct prompt.
    \item Ask the model for a second-pass \emph{verification decision} that outputs either \texttt{execute} or \texttt{clarify}, together with a short reason and a list of missing fields.
\end{enumerate}

This method is intentionally simple. It does not use fine-tuning, preference optimisation, reinforcement learning, or external retrieved knowledge. The controller only uses information already available at inference time: the user request, dialogue history, tool schema, and the model's own draft action.

\subsection{Draft Inspection and Verification}
The controller still performs a lightweight schema-aware inspection before the verifier decision is applied. In the local implementation, it:
\begin{itemize}
    \item parses the first draft \texttt{<tool\_call>} block emitted by the model;
    \item checks whether the predicted tool exists in the available schema;
    \item extracts the required fields for that tool;
    \item marks a call as under-specified if any required field is missing or filled by an empty or placeholder-like value.
\end{itemize}

A small set of ambiguity markers is still retained as a weak prior for obviously vague requests, but the main trigger is the model's own verifier decision, conditioned on the draft action and the tool schema. This makes the method stronger than a pure keyword heuristic while still remaining lightweight and reproducible.

\subsection{Escalation Policy}
The final policy used in the main comparison is \textbf{Escalation}. It preserves the direct baseline on normal cases and only triggers clarification when the draft inspection and verifier together indicate that the request is under-specified. In other words:
\begin{itemize}
    \item if the draft call is accepted by the verifier, keep the direct action;
    \item if the draft call is under-specified, ask one clarification question instead of executing the call.
\end{itemize}

This framing matters because the project is not trying to show that ``asking more questions'' is always good. The intended result is narrower: a small amount of additional interaction should be spent selectively on the cases where the direct baseline is weakest.

\subsection{Optional Repair Extension}
The repository still contains a one-step repair scaffold, but this is now treated as an optional extension rather than the core contribution. The main coursework claim should stand even without the repair experiment. This reduces engineering risk and keeps the project aligned with the strongest empirical signal currently available, namely the missing-parameter gap in the direct baseline.

\subsection{Budget Definition}
The primary budget is a \textbf{fixed interaction budget}. For the main experiments, the controller is allowed at most one additional model interaction beyond the initial draft generation. In practice:
\begin{itemize}
    \item Direct uses one model response;
    \item Clarify-Guard uses a direct draft plus, only when triggered, one clarification response;
    \item Escalation uses the same budget but spends the second interaction only on under-specified cases.
\end{itemize}

Latency and token usage are reported as secondary cost metrics, since an intervention that improves accuracy only by doubling cost would be much less convincing.

\section{Experiments}
\subsection{Research Questions}
The final experiments are organised around three core questions:
\begin{itemize}
    \item \textbf{RQ1}: How much worse is the direct baseline on under-specified multi-turn requests than on normal multi-turn requests?
    \item \textbf{RQ2}: Can Clarify-Guard reduce harmful tool calls on \texttt{multi\_turn\_miss\_param}?
    \item \textbf{RQ3}: What robustness--latency trade-off does Escalation introduce, and how much over-clarification does it cause on turns that should execute directly?
\end{itemize}

\subsection{System Variants}
The main comparison is intentionally small and controlled, as shown in Table~\ref{tab:systems}.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{0.19\linewidth}p{0.72\linewidth}}
\toprule
System & Description \\
\midrule
Direct & Official BFCL direct setting with no extra controller. \\
Clarify-Guard & Direct draft generation followed by a verifier decision that either executes the draft call or replaces it with one clarification question. \\
Escalation & Keep the direct action on normal cases, but trigger the verifier-based Clarify-Guard only when the draft call appears under-specified. \\
Repair (optional) & Secondary extension using one repair attempt after a failed direct call. \\
\bottomrule
\end{tabular}
\caption{System variants studied in this report. Repair is retained only as a secondary scaffold and is not part of the core claim.}
\label{tab:systems}
\end{table}

\subsection{Metrics}
The primary metric is the official BFCL subset accuracy. However, BFCL alone is not sufficient to explain why the controller helps or fails. We therefore add a small set of diagnostic metrics for \texttt{multi\_turn\_miss\_param}:
\begin{itemize}
    \item \textbf{Harmful-call rate on empty-ground-truth turns}: the fraction of turns whose official ground truth contains no tool call but the model still emits one;
    \item \textbf{No-call rate on empty-ground-truth turns}: the complementary rate, which is useful for showing whether the controller is actually suppressing bad actions;
    \item \textbf{No-call rate on required turns}: how often the controller suppresses a call even though the exported BFCL turn expects a tool action;
    \item \textbf{Clarification trigger rate}: how often the guard intervenes;
    \item \textbf{Mean latency and completion tokens}: the runtime cost of the intervention.
\end{itemize}

These extra metrics are important because the official BFCL score is binary at the example level. A clarification guard may improve behaviour by preventing locally harmful calls even when the overall BFCL score changes only modestly.

\subsection{Pilot Results and Interpretation}
The direct baseline is already available, and the turn-level guard pilot has also been run on 200 exported \texttt{multi\_turn\_miss\_param} turns. The resulting story is simple and coherent:
\begin{enumerate}
    \item \texttt{multi\_turn\_miss\_param} is clearly harder than \texttt{multi\_turn\_base} for the direct baseline;
    \item Clarify-Guard strongly reduces harmful premature tool calls on under-specified turns;
    \item this robustness gain is offset by substantial over-clarification and higher latency.
\end{enumerate}

The current pilot comparison is shown in Table~\ref{tab:main-results}. It should be interpreted as a turn-level diagnostic experiment rather than a replacement for the official BFCL subset accuracy.

\begin{table}[t]
\centering
\small
\begin{tabular}{lcccc}
\toprule
Method & Harmful Call Rate & No-Call on Required & Avg. Latency & Avg. Tokens \\
\midrule
Direct & 66.67\% & 22.37\% & 0.52s & 28.93 \\
Clarify-Guard & 14.58\% & 59.21\% & 0.93s & 61.11 \\
Escalation & 18.75\% & 59.21\% & 0.93s & 60.71 \\
\bottomrule
\end{tabular}
\caption{Turn-level 200-example pilot on exported \texttt{multi\_turn\_miss\_param} turns. Harmful-call rate is measured on turns whose exported ground truth expects clarification; no-call rate on required turns measures over-clarification.}
\label{tab:main-results}
\end{table}

\subsection{Discussion}
The main positive result is large and easy to interpret. On the exported \texttt{multi\_turn\_miss\_param} pilot, Clarify-Guard reduces harmful-call rate on empty-ground-truth turns from 66.67\% to 14.58\%, a drop of 52.09 percentage points. Escalation is slightly weaker on the same slice, reducing the same metric to 18.75\%. This confirms that the direct baseline often calls tools too eagerly when the request is incomplete, and that a second-pass verifier can suppress many of these mistakes.

The negative result is equally important. The same guard raises the no-call rate on required turns from 22.37\% to 59.21\%, while average latency increases from about 0.52s to about 0.93s and average completion length more than doubles. In other words, the verifier is useful but over-conservative. It often treats multi-step but still executable requests as requiring clarification. This is exactly the kind of mixed result that a coursework report can usefully analyse: robustness improves on one failure mode, but only at a substantial cost in false clarifications and runtime.

The pilot also explains why Escalation does not beat Clarify-Guard here. The exported pilot contains only \texttt{multi\_turn\_miss\_param} turns, so there are few opportunities for Escalation to preserve normal-case behaviour. In this setting, Escalation behaves very similarly to Clarify-Guard but with a small amount of extra repair overhead. For that reason, the strongest current result should be reported as the Clarify-Guard pilot, while Escalation should be treated as a secondary comparison rather than the main contribution.

\subsection{Limitations}
There are three important limitations. First, the official BFCL direct baseline uses the full benchmark evaluator, but the guard experiments use exported turn-level pilots rather than a full end-to-end BFCL rerun. This makes the guard analysis more diagnostic than leaderboard-style. Second, the study uses only one small open model and one local serving configuration, so the conclusions should be interpreted as model-specific rather than universal. Third, the verifier prompt is evidently not yet well calibrated: the over-clarification rate on required turns is high, and the current work should therefore be read as a robustness--cost trade-off study rather than a claim of overall superiority.

\section{Related Work}
\subsection{Tool Use and Agentic LLMs}
Toolformer and ReAct established the basic idea that language models can interleave generation with external actions \citep{schick2023toolformer,yao2023react}. Reflexion later showed that verbal feedback and self-improvement loops can support more robust agentic behaviour without immediate weight updates \citep{shinn2023reflexion}. These works motivate the broader setting but do not directly answer the narrow question asked in this project: whether a very lightweight controller can improve robustness for a small open model on BFCL.

\subsection{BFCL and Tool-Calling Evaluation}
BFCL is the most relevant benchmark because it evaluates structured function calling across a range of realistic conditions, including stateful multi-turn settings \citep{patil2025bfcl}. The benchmark is especially useful here because it separates normal multi-turn cases from more difficult missing-parameter and memory cases. That separation allows the project to make a much cleaner claim than a single overall score would permit.

\subsection{Clarification and Failure Handling}
Recent work has studied clarification, parameter checking, and failure diagnosis from several directions. FAIL-TaLMs highlights missing information and unavailable tools as common sources of failure \citep{trevino2025failtalms}. STYLE and AskToAct study when to ask clarification questions \citep{chen2024style,zhang2025asktoact}. HiTEC introduces structured parameter-checking for tool use \citep{cui2025hitec}, while CRITICTOOL focuses on critique and correction of erroneous tool-calling behaviour \citep{huang2025critictool}. This coursework is positioned relative to that literature as a small, focused empirical study rather than a new general-purpose method. The intended contribution is a controlled BFCL-v4 analysis of a fixed-budget clarification guard for a 4B open model.

\section{Conclusions}
The project is now tightly scoped around one concrete failure mode: under-specified multi-turn tool calls on BFCL. The current direct baseline already shows that this slice is substantially harder than the normal multi-turn setting for \PrimaryModel{}, which gives the project a clear empirical target. The proposed intervention is intentionally lightweight: a verifier-based clarification guard that suppresses incomplete draft calls, and an escalation policy that uses the guard only when necessary. The current 200-turn pilot already reveals the main trade-off. The guard sharply reduces harmful premature tool calls on under-specified turns, but it also over-clarifies many turns that should have executed directly. This is still a useful coursework result. It shows that better inference control can improve one robustness metric while harming another, and it motivates a careful discussion of when clarification is genuinely useful and when it becomes over-conservative. The final version should therefore present the method as a mixed-result robustness study grounded in both the official BFCL baseline and the new turn-level diagnostics.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
